Spatial phylogenetics of Japanese ferns: Patterns, processes, and implications for conservation

### 2021-12-24 ms/manuscript.Rmd

---
title: "Spatial phylogenetics of Japanese ferns: Patterns, processes, and implications for conservation"
mainfont: Roboto-Regular.ttf
fontsize: 11pt
mainfontoptions: 
  - BoldFont=Roboto-Bold.ttf
  - ItalicFont=Roboto-Italic.ttf
  - BoldItalicFont=Roboto-BoldItalic.ttf
output: 
  bookdown::word_document2:
    pandoc_args: [ "--csl", "american-journal-of-botany.csl"]
    toc: no
    number_sections: no
    reference_docx: ajb_template.docx
  bookdown::pdf_document2:
    citation_package: default
    toc: no
    number_sections: no
    fig_caption: no
    latex_engine: xelatex
    pandoc_args: [ "--csl", "american-journal-of-botany.csl"]
params:
  doc_type: doc
editor_options: 
  chunk_output_type: console
---

@@

Joel H. Nitta^1,5^, Brent D. Mishler^2^, Wataru Iwasaki^1,3^, and Atsushi Ebihara^4^

@@

^5^Author for correspondence: joelnitta@gmail.com

Manuscript received _______; revision accepted _______.

@@

**Abstract**

@@

@@

However, a comprehensive understanding of the relationships between richness and other metrics requires densely sampled data, and such datasets are rare on the regional (country) scale. 
The ferns of Japan are excellent model system because they have been the target of intense botanical interest for several decades and are densely sampled [reviewed in @Ebihara2019b]. 
One particularly valuable characteristic of this flora is the availability of data on reproductive mode (known for `r repro_count$known$total` of `r repro_count$known$total + repro_count$unknown$total` native pteridophyte taxa (excluding nothotaxa) [`r repro_count$known$percent`]), which is likely to affect genetic diversity [@Bengtsson2003], and thereby higher-level biodiversity [@Krueger-Hadfield2019].

@@

# Materials and methods

@@

## Study Site

@@

## Occurrence data

We used a list of fern specimens deposited at TNS [herbarium codes follow @Thiers2021]. 
We chose to use this over other sources (e.g., GBIF), because of its high taxonomic quality: all identifications are standardized to a common taxonomy, the Japan Green List v. 1.01 (http://www.rdplants.org/gl/). 
Furthermore, the overall sampling completeness of this dataset is extremely high (`r s_figure("sampling-curve")` in Appendix S1; see Supplemental Data with this article). 
The original list comprises `r occ_sum$unfiltered$n_specimens$value %>% scales::number(big.mark = ",")` specimens representing `r occ_sum$unfiltered$n_taxa$value` terminal taxa at the rank of species, subspecies, form, or variety (hereafter referred to as "taxa" unless otherwise indicated), excluding non-native taxa and nothotaxa.
Latitude and longitude of each occurrence record were projected into the Japanese Geodetic Datum 2000 (JGD 2000), which was also used for all further geographic analyses.
We filtered out any occurrences not within a map of standard grid squares defined by the Statistics Bureau of Japan (http://www.stat.go.jp/english/data/mesh/05.html) and removed duplicate collections (`r occ_sum$filtered$n_specimens$value %>% scales::number(big.mark = ",")` specimens, `r occ_sum$filtered$n_taxa$value` taxa after filtering).
The R package `r pack("sf")` was used for all GIS analyses [@Pebesma2018].

@@

## Morphological trait data

@@

## Reproductive mode data

@@

## Environmental data

@@

## Phylogenetic analysis

@@

## Phylogenetic signal

@@

## Analysis of biodiversity

@@

## Spatial models

@@

## Analysis of bioregions

@@

## Assessment of conservation status and threats

@@

# Results

## Datasets

@@

The phylogeny included `r ape::Ntip(japan_fern_tree)` taxa (`r phylogeny_sampled_percent` of the total number of native ferns excluding nothotaxa [`r total_fern_taxa` taxa]) representing `r n_fern_phylo_genera` genera and `r n_fern_phylo_families` families. 
The topology was in general agreement with other recent, large plastid fern phylogenies [@Schuettpelz2007; @Lehtonen2011; @Testo2016a] (`r s_figure("tree")`). 
Ferns were recovered as monophyletic, as were all families and genera except for *`r non_mono_genus`* [families follow @PteridophytePhylogenyGroupI2016]. 
The age of the fern clade is estimated to be `r phylogeny_age` ma.

@@

## Observed diversity

@@

## Randomization tests for phylogenetic and functional diversity

@@

## Relationships between biodiversity and climate or reproductive mode

@@

## Categorical analysis of neo- and paleo-endemism

@@

## Bioregions

@@

Taxonomic bioregions 3 and 4 have higher SES values of PD, RPD, FD, and RFD than bioregions 1 and 2 (`r figure("div-by-tax-region")`). 
Bioregions 3 and 4 also have consistently high PE *P*-scores, whereas bioregions 1 and 2 have a much wider variation. 
Bioregion 2 has lower RPD and RFD than bioregion 1. 
Comparison of biodiversity metrics across phylogenetic bioregions was similar, except that phylogenetic bioregion 3 mostly corresponds to taxonomic bioregions 3 and 4 combined.

## Conservation status and threats

@@

# Discussion

## Distribution and drivers of biodiversity

@@

## Conservation status and threats

@@

# Conclusions

@@

Our study integrates diverse, thoroughly collected datasets to characterize the spatial biodiversity of Japanese ferns and gain insight into the processes generating these patterns. We conclude by summarizing some caveats that must be kept in mind when interpreting our results, while also suggesting avenues for future research.

First, the data include specimens collected over a wide time period, so they do not necessarily represent the current distribution of ferns in Japan. 
Furthermore, the collections are not evenly collected throughout the country, but rather some areas have many more specimens than others (`r s_figure("abundance")`). 
Indeed, one promising avenue of research may be to take advantage of the most densely sampled areas to investigate changes in abundance over time.

@@

One element of biodiversity that we did not consider here is nothotaxa (hybrids). 
The Japanese fern flora includes `r n_hybrid_fern_taxa` nothotaxa. 
Although hybrids are often thought of as evolutionary "dead-ends", in some cases they are able to interbreed with diploids and thus contribute to diversification [@Barrington1989]. 
Future studies should consider the distribution of hybrids in Japan and their possible effects on biodiversity.
Methods for calculating PD would need to be modified to take into account multiple lineages contributing to one terminal taxon.

@@

# Define nicely formatted terms for printing variable names
resp_var_lookup <- tribble(
  ~resp_var, ~resp_var_pretty,
  "fd_obs_z", "SES of FD",
  "pd_obs_z", "SES of PD",
  "pe_obs_p_upper", "PE *P*-value",
  "rfd_obs_z", "SES of RFD",
  "richness", "Richness",
  "rpd_obs_z", "SES of RPD",
)

@@

  ggplot(plot_data, aes(x = t_value, y = term_print, fill = signif)) +
    geom_vline(xintercept = 0, linetype = "dotted") +
    geom_col() +
    # scale_x_continuous(n.breaks = 4) +
    scale_x_continuous() +
    scale_fill_manual(
      values = c("TRUE" = "grey20", "FALSE" = "grey70"),
      labels = c("TRUE" = "*P* < 0.05", "FALSE" = "*P* > 0.05")) +
    labs(
      subtitle = glue("{unique(plot_data$resp_var_print)},\n{unique(plot_data$model_type_print)}"),
      fill = "Significance") +
    theme_bw(base_size = 9) +
    theme(
      axis.title = element_blank(),
      axis.text.y = element_markdown(),
      legend.text = element_markdown(),
      panel.grid.minor = element_blank(),
      legend.position = "none")
}

@@

e <- ggplot(data, aes(x = taxonomic_cluster, y = pe_obs_p_upper)) +
  geom_jitter(aes(color = taxonomic_cluster)) +
  scale_color_manual(values = bioregion_cols, breaks = 1:4) +
  geom_boxplot(color = "grey80", fill = "transparent", outlier.shape = NA, alpha = 0.25) +
  labs(
    y = "PE *P*-score",
    x = "Cluster",
    color = "Bioregion"
  ) +
  theme_bw() +
  theme(
    axis.title.y = element_markdown(),
    panel.grid.minor.x = element_blank(),
    panel.grid.major.x = element_blank(),
    panel.grid.minor.y = element_blank(),
  )

@@

**`r figure("conserv-status")`**. Conservation status and threat due to herbivory by deer in areas with high levels of biodiversity for ferns of Japan. (A) Percent of land area with medium or high protected status for grid-cells with significantly high biodiversity by protection status. (B) Percent of land area occupied by deer for grid-cells with significantly high biodiversity by deer range data source (1978 survey, 2003 survey, or range estimated from model based on 2003 survey data). Biodiversity metrics include taxon richness, phylogenetic diversity (PD), functional diversity (FD), and phylogenetic endemism (PE). Significance of PD, FD, and PE assessed by a one-tailed test comparing observed values to a null distribution of 999 random values; for richness, grid-cells in the top 5% considered highly diverse (see Materials and Methods). For (A), vertical lines indicate total percent area protected across Japan (baseline protection rate). For (B), vertical lines indicate total percentage of land area occupied by deer across Japan (baseline threat rate).

>>>>

---
title: "Spatial phylogenetics of Japanese ferns: Patterns, processes, and implications for conservation"
mainfont: Roboto-Regular.ttf
fontsize: 11pt
mainfontoptions: 
  - BoldFont=Roboto-Bold.ttf
  - ItalicFont=Roboto-Italic.ttf
  - BoldItalicFont=Roboto-BoldItalic.ttf
output: 
  bookdown::word_document2:
    pandoc_args: [ "--csl", "apa-6th-edition.csl"]
    toc: no
    number_sections: no
    reference_docx: template.docx
  bookdown::pdf_document2:
    citation_package: default
    toc: no
    number_sections: no
    fig_caption: no
    latex_engine: xelatex
    pandoc_args: [ "--csl", "apa-6th-edition.csl"]
params:
  doc_type: doc
editor_options: 
  chunk_output_type: console
---

@@

Joel H. Nitta^1^, Brent D. Mishler^2^, Wataru Iwasaki^1,3^, and Atsushi Ebihara^4^

@@

Author for correspondence:  
*Joel H. Nitta*  
*Email: joelnitta@gmail.com*

## Acknowledgements

Members of the Iwasaki Lab at the University of Tokyo provided comments that improved the manuscript.
Shawn Laffan provided advice about CANAPE.
This study was supported by JSPS KAKENHI Grant Number 16H06279.

@@

## Abstract

@@

## Introduction

@@

However, a comprehensive understanding of the relationships between richness and other metrics requires densely sampled data, and such datasets are rare on the regional (country) scale. 
The ferns of Japan are excellent model system because they have been the target of intense botanical interest for several decades and are densely sampled [reviewed in @Ebihara2019b]. 
One particularly valuable characteristic of this flora is the availability of data on reproductive mode (known for `r repro_count$known$total` of `r repro_count$known$total + repro_count$unknown$total` native pteridophyte taxa excluding hybrids [`r repro_count$known$percent`]), which is likely to affect genetic diversity [@Bengtsson2003], and thereby higher-level biodiversity [@Krueger-Hadfield2019].

@@

## Materials and Methods

@@

### Study Site

@@

### Occurrence data

We used a list of all native, non-hybrid fern specimens deposited at TNS [herbarium codes follow @Thiers2021]. 
We chose to use this over other sources (e.g., GBIF), because of its high taxonomic quality: all identifications are standardized to a common taxonomy, the Japan Green List v. 1.01 (http://www.rdplants.org/gl/). 
Furthermore, the overall sampling completeness of this dataset is extremely high (`r s_figure("sampling-curve")` in Appendix S1; see Supplemental Data with this article). 
The original list comprises `r occ_sum$unfiltered$n_specimens$value %>% scales::number(big.mark = ",")` specimens representing `r occ_sum$unfiltered$n_taxa$value` terminal taxa at the rank of species, subspecies, form, or variety (hereafter referred to as "taxa" unless otherwise indicated), excluding non-native taxa and nothotaxa.
Latitude and longitude of each occurrence record were projected into the Japanese Geodetic Datum 2000 (JGD 2000), which was also used for all further geographic analyses.
We filtered out any occurrences not within a map of standard grid squares defined by the Statistics Bureau of Japan (http://www.stat.go.jp/english/data/mesh/05.html) and removed duplicate collections (`r occ_sum$filtered$n_specimens$value %>% scales::number(big.mark = ",")` specimens, `r occ_sum$filtered$n_taxa$value` taxa after filtering).
The R package `r pack("sf")` was used for all GIS analyses [@Pebesma2018].

@@

### Morphological trait data

@@

### Reproductive mode data

@@

### Environmental data

@@

### Phylogenetic analysis

@@

### Phylogenetic signal

@@

### Analysis of biodiversity

@@

### Spatial models

@@

### Analysis of bioregions

@@

### Assessment of conservation status and threats

@@

## Results

### Datasets

@@

The phylogeny included `r ape::Ntip(japan_fern_tree)` taxa (`r phylogeny_sampled_percent` of the total number of native, non-hybrid ferns [`r total_fern_taxa` taxa]) representing `r n_fern_phylo_genera` genera and `r n_fern_phylo_families` families. 
The topology was in general agreement with other recent, large plastid fern phylogenies [@Schuettpelz2007; @Lehtonen2011; @Testo2016a] (`r s_figure("tree")`). 
Ferns were recovered as monophyletic, as were all families and genera except for *`r non_mono_genus`* [families follow @PteridophytePhylogenyGroupI2016]. 
The age of the fern clade is estimated to be `r phylogeny_age` ma.

@@

### Observed diversity

@@

### Randomization tests for phylogenetic and functional diversity

@@

### Relationships between biodiversity and climate or reproductive mode

@@

### Categorical analysis of neo- and paleo-endemism

@@

### Bioregions

@@

Taxonomic bioregions 3 and 4 have higher SES values of PD, RPD, FD, and RFD than bioregions 1 and 2 (`r figure("div-by-tax-region")`). 
Bioregions 3 and 4 also have consistently high PE *p*-scores, whereas bioregions 1 and 2 have a much wider variation. 
Bioregion 2 has lower RPD and RFD than bioregion 1. 
Comparison of biodiversity metrics across phylogenetic bioregions was similar, except that phylogenetic bioregion 3 mostly corresponds to taxonomic bioregions 3 and 4 combined.

### Conservation status and threats

@@

## Discussion

### Distribution and drivers of biodiversity

@@

### Conservation status and threats

@@

### Limits of the study and future outlook

@@

There are a number of caveats that must be kept in mind when interpreting these results, but these also suggest avenues for future research. 
First, the data include specimens collected over a wide time period, so they do not necessarily represent the current distribution of ferns in Japan. 
Furthermore, the collections are not evenly collected throughout the country, but rather some areas have many more specimens than others (`r s_figure("abundance")`). 
Indeed, one promising avenue of research may be to take advantage of the most densely sampled areas to investigate changes in abundance over time.

@@

One element of biodiversity that we did not consider here is hybrids. 
The Japanese fern flora includes `r n_hybrid_fern_taxa` hybrid taxa. 
Although hybrids are often thought of as evolutionary "dead-ends", in some cases they are able to interbreed with diploids and thus contribute to diversification [@Barrington1989]. 
Future studies should consider the distribution of hybrids in Japan and their possible effects on biodiversity.
Methods for calculating PD would need to be modified to take into account multiple lineages contributing to one terminal taxon.

@@

# Define nicely formatted terms for printing variable names
resp_var_lookup <- tribble(
  ~resp_var, ~resp_var_pretty,
  "fd_obs_z", "SES of FD",
  "pd_obs_z", "SES of PD",
  "pe_obs_p_upper", "PE *p*-value",
  "rfd_obs_z", "SES of RFD",
  "richness", "Richness",
  "rpd_obs_z", "SES of RPD",
)

@@

  ggplot(plot_data, aes(x = t_value, y = term_print, fill = signif)) +
    geom_vline(xintercept = 0, linetype = "dotted") +
    geom_col() +
    # scale_x_continuous(n.breaks = 4) +
    scale_x_continuous() +
    scale_fill_manual(
      values = c("TRUE" = "grey20", "FALSE" = "grey70"),
      labels = c("TRUE" = "*p* < 0.05", "FALSE" = "*p* > 0.05")) +
    labs(
      subtitle = glue("{unique(plot_data$resp_var_print)},\n{unique(plot_data$model_type_print)}"),
      fill = "Significance") +
    theme_bw(base_size = 9) +
    theme(
      axis.title = element_blank(),
      axis.text.y = element_markdown(),
      legend.text = element_markdown(),
      panel.grid.minor = element_blank(),
      legend.position = "none")
}

@@

e <- ggplot(data, aes(x = taxonomic_cluster, y = pe_obs_p_upper)) +
  geom_jitter(aes(color = taxonomic_cluster)) +
  scale_color_manual(values = bioregion_cols, breaks = 1:4) +
  geom_boxplot(color = "grey80", fill = "transparent", outlier.shape = NA, alpha = 0.25) +
  labs(
    y = "PE *p*-score",
    x = "Cluster",
    color = "Bioregion"
  ) +
  theme_bw() +
  theme(
    axis.title.y = element_markdown(),
    panel.grid.minor.x = element_blank(),
    panel.grid.major.x = element_blank(),
    panel.grid.minor.y = element_blank(),
  )

@@

**`r figure("conserv-status")`**. Conservation status and threat due to herbivory by deer in areas with high levels of biodiversity for ferns of Japan. (a) Percent of land area with medium or high protected status for grid-cells with significantly high biodiversity by protection status. (b) Percent of land area occupied by deer for grid-cells with significantly high biodiversity by deer range data source (1978 survey, 2003 survey, or range estimated from model based on 2003 survey data). Biodiversity metrics include taxon richness, phylogenetic diversity (PD), functional diversity (FD), and phylogenetic endemism (PE). Significance of PD, FD, and PE assessed by a one-tailed test comparing observed values to a null distribution of 999 random values; for richness, grid-cells in the top 5% considered highly diverse (see Materials and Methods). For (a), vertical lines indicate total percent area protected across Japan (baseline protection rate). For (b), vertical lines indicate total percentage of land area occupied by deer across Japan (baseline threat rate).

`r pagebreak_doc()`

```{r japan-map-doc, fig.cap = "", fig.width = 5, fig.height = 5, eval = params$doc_type == "doc", results = "asis"}
# Print plot if rendering to doc on initial submission
japan_map_plot
cat("\n\n", figure("japan-map"))
```

`r pagebreak_doc()`

```{r raw-diversity-doc, fig.cap = "", fig.width = 7, fig.height = 4, eval = params$doc_type == "doc", results = "asis"}
raw_diversity_plot
cat("\n\n", figure("raw-diversity"))
```

`r pagebreak_doc()`

```{r rand-diversity-doc, fig.cap = "", fig.width = 7, fig.height = 9, eval = params$doc_type == "doc", results = "asis"}
rand_diversity_plot
cat("\n\n", figure("rand-diversity"))
```

`r pagebreak_doc()`

```{r coeff-doc, fig.cap = "", fig.width = 7, fig.height = 5, eval = params$doc_type == "doc", results = "asis"}
coeff_plot
cat("\n\n", figure("coeff"))
```

`r pagebreak_doc()`

```{r modelfit-doc, fig.cap = "", fig.width = 7, fig.height = 5, eval = params$doc_type == "doc", results = "asis"}
modelfit_plot
cat("\n\n", figure("modelfit"))
```

`r pagebreak_doc()`

```{r endemism-doc, fig.cap = "", fig.width = 5, fig.height = 5, eval = params$doc_type == "doc", results = "asis"}
endemism_plot
cat("\n\n", figure("endemism"))
```

`r pagebreak_doc()`

```{r bioregions-doc, fig.cap = "", fig.width = 7, fig.height = 5, eval = params$doc_type == "doc", results = "asis"}
bioregions_plot
cat("\n\n", figure("bioregions"))
```

`r pagebreak_doc()`

```{r div-by-tax-region-doc, fig.cap = "", fig.width = 7, fig.height = 7, eval = params$doc_type == "doc", results = "asis"}
div_by_tax_region_plot
cat("\n\n", figure("div-by-tax-region"))
```

`r pagebreak_doc()`

```{r conserv-status-doc, fig.cap = "", fig.width = 7, fig.height = 4, eval = params$doc_type == "doc", results = "asis"}
conserv_status_plot
cat("\n\n", figure("conserv-status"))
```

`r pagebreak_doc()`

## Data availability statement

Code to replicate all analyses, figures, and this manuscript are available at https://github.com/joelnitta/japan_ferns_spatial_phy. 
A Docker image to run the code is available at https://hub.docker.com/r/joelnitta/japan_ferns_spatial_phy.
Data files needed to run the analysis and selected results files are available from Dryad at http://doi.org/FIXME:ADD_DOI_HERE_WHEN_ASSIGNED.

## References
<!-- Insert <div id="refs"></div> to place the refs here instead of at the end -->
<div id="refs"></div>

## Biosketch

Joel H. Nitta's research focuses on patterns and processes of diversification in ferns. For more information, see https://joelnitta.com.

Author contributions: J.H.N., B.M., and A.E. conceived the ideas; A.E. provided the data; J.H.N. analyzed the data; W.I provided resources; J.H.N. wrote the manuscript with input from all co-authors.
